---
title: "Schelling Agent-Based Segregation Model with variations"
author: "Jon Green"
date: "June 12, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(1212)
```

I replicated Schelling's segregation model (using David Zimmerman's example at https://www.r-bloggers.com/agent-based-modelling-with-data-table-or-how-to-model-urban-migration-with-r/ as a starting point) using data.table() and ggplot() in order to organize, update, and visualize how agents were interacting in the wrapped universe I specified.

The first function initiates the Schelling universe, taking a specified grid size, number of races, percent of empty cells, and percent of agents of the dominant race, the second function plots the Schelling universe, and the third function determines how the agents interact while iterating through multiple rounds of movement. This third function also stores the percentage of agents who are unsatisfied in each iteration, putting a vector of percentages in the global environment that can be plotted.

```{r}
initiateSchelling_wrapped <- function(dimensions = c(10, 10), n_races = 4, perc_empty = 0.2, perc_maj = .75){
  require(data.table)
  dims <<- dimensions
  races <<- colours()[1:n_races] # create "races" based on colors
  
  # get number of homes (equal to number of cells)
  n_homes = prod(dimensions)
  # calculates the number of agents (cells minus the percent of empty homes)
  count_agents <- floor(n_homes * (1 - perc_empty))
  
  # the characteristics that a home can have -- either empty or a particular race
  races <- c("empty", races)
  # the probabilities of a home having each characteristic: 
  #   probability of a cell being empty, of being occupied by the dominant race conditional on not being empty, 
  #   and of being in a minority race conditional on not being empty
  probabilities <- c(perc_empty, 
                     (1-perc_empty)*perc_maj, 
                     rep((1 - perc_empty - (1-perc_empty)*perc_maj)/(n_races-1), times = n_races-1)
  )
  
  # cast the schelling data.table in the global environment
  schelling <<- data.table(id = 1:prod(dimensions),
                           x = rep(1:dimensions[1], 
                                   dimensions[2]),
                           y = rep(1:dimensions[2], 
                                   each = dimensions[1]),
                           race = sample(x = races, 
                                         size = n_homes, 
                                         prob = probabilities,
                                         replace = TRUE),
                           unsatisfied = rep(NA, prod(dimensions))) # leave this column empty to fill later
}
```

The second function plots the grid.
```{r}
plotSchelling <- function(title){
  require(data.table)
  require(ggplot2)
  require(RColorBrewer)
  # get races to get the right number of colors
  races <- unique(schelling$race)
  
  # find the dimensions of the grid to get the best dot size
  dims <- c(max(schelling$x), max(schelling$y))
  
  # create colors
  # check if there are 3 or fewer races, 
  # RColorBrewer otherwise
  if (length(races) <= 3) {
    colors <- brewer.pal(3, "Dark2")
  } else {
    colors <- brewer.pal(length(races), "Dark2")
  }
  
  # plot the graph  
  p <- ggplot(data = schelling[race != "empty"], 
              aes(x = x, y = y, color = race)) + 
    # resize dots to grid
    geom_point(size = 100/sqrt(prod(dims))) +  
    scale_colour_manual(values = colors) + 
    # theme: mostly blank
    theme_bw() + 
    theme(axis.line = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          legend.position = "none",
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          plot.background = element_blank(),
          plot.title = element_text(lineheight=3, 
                                    face="bold", 
                                    color="black", 
                                    size=14)) +
    # fix axes to avoid distortion in the picture
    coord_fixed(ratio = 1) + 
    # add the title
    ggtitle(title)
  
  return(p)
}
```

In the first version of the model, each agent looks at their eight surrounding neighbors (agents at the edges and corners look to the other side of the grid such that the universe wraps), and if the percent of like neighbors is below their preference threshold (thresholds vary depending on whether the agent is a member of the majority race -- default is for the majority race to prefer a higher percentage of like neighbors) they move to a cell that is either empty or occupied by another unsatisfied agent.

```{r}
iterate_wrapped <- function(n = 10, dom_sim_threshold = .9, min_sim_threshold = .1){
  require(data.table)
  # subfunction that checks for a given x and y value if the agent is unsatisfied (returns TRUE or FALSE)
  is.unsatisfied_wrapped <- function(x_value, y_value, 
                                     dom_sim_threshold = dom_sim_threshold,
                                     min_sim_threshold = min_sim_threshold){
    # gets the race and id for the agent
    cur_race <- schelling[x == x_value & y == y_value, race]
    cur_id <- schelling[x == x_value & y == y_value, id]
    
    #sub-subfunction to identify agent's neighbors
    select.neighbors <- function(dimensions){
      
      #Identify coordinates of neighbors (adjusts wrapping based on position of agent)
      neighbors <- as.data.frame(matrix(NA, nrow = 8, ncol = 2))
      if(x_value == 1 & y_value == 1){
        neighbors[1,] <- schelling[x == dimensions[1] & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == dimensions[1] & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == dimensions[1] & y == dimensions[2], 2:3]
        neighbors[7,] <- schelling[x == x_value & y == dimensions[2], 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == dimensions[2], 2:3]
      }
      if(x_value == 1 & y_value %in% c(2:(dimensions[2]-1))){
        neighbors[1,] <- schelling[x == dimensions[1] & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == dimensions[1] & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == dimensions[1] & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == y_value-1, 2:3]
      }
      if(x_value == 1 & y_value == dimensions[2]){
        neighbors[1,] <- schelling[x == dimensions[1] & y == 1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == 1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == 1, 2:3]
        neighbors[4,] <- schelling[x == dimensions[1] & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == dimensions[1] & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == y_value-1, 2:3]
      }
      if(x_value %in% c(2:(dimensions[1]-1)) & y_value == 1){
        neighbors[1,] <- schelling[x == x_value-1 & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == dimensions[2], 2:3]
        neighbors[7,] <- schelling[x == x_value & y == dimensions[2], 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == dimensions[2], 2:3]
      }
      if(x_value %in% c(2:(dimensions[1]-1)) & y_value %in% c(2:(dimensions[2]-1))){
        neighbors[1,] <- schelling[x == x_value-1 & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == y_value-1, 2:3] 
      }
      if(x_value %in% c(2:(dimensions[1]-1)) & y_value == dimensions[2]){
        neighbors[1,] <- schelling[x == x_value-1 & y == 1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == 1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == 1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == y_value-1, 2:3]
      }
      if(x_value == dimensions[1] & y_value == 1){
        neighbors[1,] <- schelling[x == x_value-1 & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == 1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == 1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == dimensions[2], 2:3]
        neighbors[7,] <- schelling[x == x_value & y == dimensions[2], 2:3]
        neighbors[8,] <- schelling[x == 1 & y == dimensions[2], 2:3]
      }
      if(x_value == dimensions[1] & y_value %in% c(2:(dimensions[2]-1))){
        neighbors[1,] <- schelling[x == x_value-1 & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == 1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == 1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == 1 & y == y_value-1, 2:3]
      }
      if(x_value == dimensions[1] & y_value == dimensions[2]){
        neighbors[1,] <- schelling[x == x_value-1 & y == 1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == 1, 2:3]
        neighbors[3,] <- schelling[x == 1 & y == 1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == 1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == 1 & y == y_value-1, 2:3]
      }
      names(neighbors) <- c("x","y")
      
      ids <- schelling[x %in% neighbors$x & y %in% neighbors$y, id]
      
      return(ids)
    }
    # checks if the home is empty
    if (cur_race == "empty"){
      return(FALSE) # empty houses can't be unsatisfied
    }else{
      neighbor.ids <- select.neighbors(dimensions = dims)
      
      # counts simliar neighbors
      count_similar <- sum(schelling[id %in% neighbor.ids, 
                                     id != cur_id &
                                       race == cur_race])
      # counts different neighbors
      count_different <- sum(schelling[id %in% neighbor.ids, 
                                       id != cur_id &
                                         race != cur_race & 
                                         race != "empty"])
      
      # calculates the ratio
      ratio <- count_similar/(count_similar + count_different)
      
      # returns TRUE if the ratio is less than or equal to the threshold
      if(cur_race == races[1]){
        return(ratio <= dom_sim_threshold)
      }else{
        return(ratio <= min_sim_threshold)
      }
    }
  }
  
  # creates a ProgressBar
  pb <- txtProgressBar(min = 0, max = 1, style = 3)
  # and tracks time
  t <- Sys.time()
  
  #sets empty vector of ratios
  ratios.unsatisfied <<- c(NULL)
  
  # iterates
  for (iterate in 1:n){
    # fills the boolean vector "unsatisfied" 
    # indicates if the household is unsatisfied 
    schelling[, unsatisfied := is.unsatisfied_wrapped(x_value = x, 
                                                      y_value = y, 
                                                      dom_sim_threshold = dom_sim_threshold,
                                                      min_sim_threshold = min_sim_threshold),
              by = 1:nrow(schelling)]
    #hacky fix for NAs, which appear to happen when the ratio exactly equals the threshold (and should therefore evaluate to satisfied)
    schelling$unsatisfied[is.na(schelling$unsatisfied)] <- FALSE
    
    ratio.unsatisfied <- sum(schelling$unsatisfied)/nrow(schelling)
    ratios.unsatisfied <<- c(ratios.unsatisfied, ratio.unsatisfied)
    
    # move unsatisfied agents to an empty house
    # find the IDs that are empty where agents can migrate to
    emptyIDs <- schelling[race == "empty", id] # finds the id of empty houses
    
    # finds the origin of the agents moving, 
    oldIDs <- schelling[(unsatisfied), id] # origin
    
    # generates new IDs for each household moving
    # note that households can move to the house of other moving agents
    # in this version, agents can (by a small chance) choose to "move" to their 
    # existing home
    newIDs <- sample(x = c(emptyIDs, oldIDs), 
                     size = length(oldIDs), 
                     replace = F) # target
    
    # cast a new data.table showing migration from origin_id to target-id
    transition <- data.table(origin = oldIDs, 
                             oldRace = schelling[id %in% oldIDs, race],
                             target = newIDs)
    
    # move agents to the new homes
    schelling[id %in% transition$origin]$race = "empty"
    schelling[id %in% transition$target]$race = transition$oldRace
    
    # orders and reset schelling
    schelling <<- schelling[order(id)]
    
    # updates the ProgressBar
    setTxtProgressBar(pb, iterate/n)
  }
  close(pb)
  timedif <- Sys.time() - t
  
  # print out statistics for the calculation time
  print(paste0("Time for calculation in seconds: ", round(timedif, 3), " or: ",
               round(n / as.numeric(timedif), 3), " iterations per second"))
  print(ratios.unsatisfied)
  plotSchelling(title = paste0("Schelling Segregation Model after ", n, " Iterations"))
}
```

In the second version of the model, everything is the same except unsatisfied agents move to NEAREST AVAILABLE cell. For each iteration, each agent checks whether the percentage of like neighbors is below their preference threshold, and moves to to nearest available cell (empty or occupied by an unsatisfied agent) if they are unsatisfied.

```{r}
iterate_wrapped_nearest <- function(n = 10, dom_sim_threshold = .9, min_sim_threshold = .1){
  require(data.table)
  # subfunction that checks for a given x and y value if the agent is 
  # unsatisfied (returns TRUE or FALSE)
  is.unsatisfied_wrapped <- function(x_value, y_value, 
                                     dom_sim_threshold = dom_sim_threshold,
                                     min_sim_threshold = min_sim_threshold){
    # gets the race and id for the agent
    cur_race <- schelling[x == x_value & y == y_value, race]
    cur_id <- schelling[x == x_value & y == y_value, id]
    
    #sub-subfunction to identify agent's neighbors
    select.neighbors <- function(dimensions){
      
      #Identify coordinates of neighbors (adjusts wrapping based on position of agent)
      neighbors <- as.data.frame(matrix(NA, nrow = 8, ncol = 2))
      if(x_value == 1 & y_value == 1){
        neighbors[1,] <- schelling[x == dimensions[1] & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == dimensions[1] & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == dimensions[1] & y == dimensions[2], 2:3]
        neighbors[7,] <- schelling[x == x_value & y == dimensions[2], 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == dimensions[2], 2:3]
      }
      if(x_value == 1 & y_value %in% c(2:(dimensions[2]-1))){
        neighbors[1,] <- schelling[x == dimensions[1] & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == dimensions[1] & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == dimensions[1] & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == y_value-1, 2:3]
      }
      if(x_value == 1 & y_value == dimensions[2]){
        neighbors[1,] <- schelling[x == dimensions[1] & y == 1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == 1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == 1, 2:3]
        neighbors[4,] <- schelling[x == dimensions[1] & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == dimensions[1] & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == y_value-1, 2:3]
      }
      if(x_value %in% c(2:(dimensions[1]-1)) & y_value == 1){
        neighbors[1,] <- schelling[x == x_value-1 & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == dimensions[2], 2:3]
        neighbors[7,] <- schelling[x == x_value & y == dimensions[2], 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == dimensions[2], 2:3]
      }
      if(x_value %in% c(2:(dimensions[1]-1)) & y_value %in% c(2:(dimensions[2]-1))){
        neighbors[1,] <- schelling[x == x_value-1 & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == y_value-1, 2:3] 
      }
      if(x_value %in% c(2:(dimensions[1]-1)) & y_value == dimensions[2]){
        neighbors[1,] <- schelling[x == x_value-1 & y == 1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == 1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == 1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == y_value-1, 2:3]
      }
      if(x_value == dimensions[1] & y_value == 1){
        neighbors[1,] <- schelling[x == x_value-1 & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == 1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == 1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == dimensions[2], 2:3]
        neighbors[7,] <- schelling[x == x_value & y == dimensions[2], 2:3]
        neighbors[8,] <- schelling[x == 1 & y == dimensions[2], 2:3]
      }
      if(x_value == dimensions[1] & y_value %in% c(2:(dimensions[2]-1))){
        neighbors[1,] <- schelling[x == x_value-1 & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == 1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == 1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == 1 & y == y_value-1, 2:3]
      }
      if(x_value == dimensions[1] & y_value == dimensions[2]){
        neighbors[1,] <- schelling[x == x_value-1 & y == 1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == 1, 2:3]
        neighbors[3,] <- schelling[x == 1 & y == 1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == 1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == 1 & y == y_value-1, 2:3]
      }
      names(neighbors) <- c("x","y")
      
      ids <- schelling[x %in% neighbors$x & y %in% neighbors$y, id]
      
      return(ids)
    }
    # checks if the home is empty
    if (cur_race == "empty"){
      return(FALSE) # empty houses can't be unsatisfied
    }else{
      neighbor.ids <- select.neighbors(dimensions = dims)
      
      # counts simliar neighbors
      count_similar <- sum(schelling[id %in% neighbor.ids, 
                                     id != cur_id &
                                       race == cur_race])
      # counts different neighbors
      count_different <- sum(schelling[id %in% neighbor.ids, 
                                       id != cur_id &
                                         race != cur_race & 
                                         race != "empty"])
      
      # calculates the ratio
      ratio <- count_similar/(count_similar + count_different)
      
      # returns TRUE if the ratio is less than or equal to the threshold
      if(cur_race == races[1]){
        return(ratio <= dom_sim_threshold)
      }else{
        return(ratio <= min_sim_threshold)
      }
    }
  }
  
  # creates a ProgressBar
  pb <- txtProgressBar(min = 0, max = 1, style = 3)
  # and tracks time
  t <- Sys.time()
  
  #sets empty vector of ratios
  ratios.unsatisfied <<- c(NULL)
  
  # iterates
  for (iterate in 1:n){
    # fills the boolean vector "unsatisfied" 
    # indicates if the household is unsatisfied 
    schelling[, unsatisfied := is.unsatisfied_wrapped(x_value = x, 
                                                      y_value = y, 
                                                      dom_sim_threshold = dom_sim_threshold,
                                                      min_sim_threshold = min_sim_threshold),
              by = 1:nrow(schelling)]
    #hacky fix for NAs, which appear to happen when the ratio exactly equals the threshold (and should therefore evaluate to satisfied)
    schelling$unsatisfied[is.na(schelling$unsatisfied)] <- FALSE
    
    ratio.unsatisfied <- sum(schelling$unsatisfied)/nrow(schelling)
    ratios.unsatisfied <<- c(ratios.unsatisfied, ratio.unsatisfied)
    
    # move unsatisfied agents to an empty house
    # find the IDs that are empty where agents can migrate to
    emptyIDs <- schelling[race == "empty", id] # finds the id of empty houses
    
    usIDs <- schelling[(unsatisfied), id] # finds the origin of the agents moving, 
    
    availIDs <<- sort(c(emptyIDs, usIDs)) # finds available IDs for each household moving
    
    #subfunction to find nearest available cell
    find.nearest <- function(cell, avails){
      distance2 <- c(NULL) #cast empty vector to fill with distances to available cells
      for(i in 1:length(avails)){
        #fill distances vector with squared distance (to keep all values positive) between current cell and each available cell
        distance2[i] <- (schelling[id == cell, x] - schelling[id == avails[i], x])^2 + (schelling[id == cell, y] - schelling[id == avails[i], y])^2
      }
      distance2 <- distance2[which(distance2 > 0)] #drop current cell
      min.id <- which.min(replace(distance2, distance2<=0, NA)) #identify id of closest available cell
      availIDs <<- avails[-which(avails == avails[min.id])] #remove that cell from list of available ids
      return(avails[min.id]) #return id
    }
    
    # cast a new data.table showing migration from origin_id to target-id
    transition <- data.table(origin = usIDs, 
                             oldRace = schelling[id %in% usIDs, race])
    transition[, target := find.nearest(cell = origin, avails = availIDs),
               by = 1:nrow(transition)]
    
    # move agents to the new homes
    schelling[id %in% transition$origin]$race = "empty"
    schelling[id %in% transition$target]$race = transition$oldRace
    
    # orders and reset schelling
    schelling <<- schelling[order(id)]
    
    # updates the ProgressBar
    setTxtProgressBar(pb, iterate/n)
  }
  close(pb)
  timedif <- Sys.time() - t
  
  # print out statistics for the calculation time
  print(paste0("Time for calculation in seconds: ", round(timedif, 3), " or: ",
               round(n / as.numeric(timedif), 3), " iterations per second"))
  print(ratios.unsatisfied)
  plotSchelling(title = paste0("Schelling Segregation Model after ", n, " Iterations"))
}
```

In the third version of the model, unsatisfied agents move to RANDOM AVAILABLE CELL WITHIN A DISTANCE LIMIT. Unsatisfied agents move to a cell sampled from the list of available (empty and occupied by an unsatisfied agent) cells that are less than a specified number of one-unit moves away from the current cell.

```{r}
iterate_wrapped_limit <- function(n = 10, dom_sim_threshold = .5, min_sim_threshold = .5, limit = 8){
  require(data.table)
  # subfunction that checks for a given x and y value if the agent is 
  # unsatisfied (returns TRUE or FALSE)
  is.unsatisfied_wrapped <- function(x_value, y_value, 
                                     dom_sim_threshold = dom_sim_threshold,
                                     min_sim_threshold = min_sim_threshold){
    # gets the race and id for the agent
    cur_race <- schelling[x == x_value & y == y_value, race]
    cur_id <- schelling[x == x_value & y == y_value, id]
    
    #sub-subfunction to identify agent's neighbors
    select.neighbors <- function(dimensions){
      
      #Identify coordinates of neighbors (adjusts wrapping based on position of agent)
      neighbors <- as.data.frame(matrix(NA, nrow = 8, ncol = 2))
      if(x_value == 1 & y_value == 1){
        neighbors[1,] <- schelling[x == dimensions[1] & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == dimensions[1] & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == dimensions[1] & y == dimensions[2], 2:3]
        neighbors[7,] <- schelling[x == x_value & y == dimensions[2], 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == dimensions[2], 2:3]
      }
      if(x_value == 1 & y_value %in% c(2:(dimensions[2]-1))){
        neighbors[1,] <- schelling[x == dimensions[1] & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == dimensions[1] & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == dimensions[1] & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == y_value-1, 2:3]
      }
      if(x_value == 1 & y_value == dimensions[2]){
        neighbors[1,] <- schelling[x == dimensions[1] & y == 1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == 1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == 1, 2:3]
        neighbors[4,] <- schelling[x == dimensions[1] & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == dimensions[1] & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == y_value-1, 2:3]
      }
      if(x_value %in% c(2:(dimensions[1]-1)) & y_value == 1){
        neighbors[1,] <- schelling[x == x_value-1 & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == dimensions[2], 2:3]
        neighbors[7,] <- schelling[x == x_value & y == dimensions[2], 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == dimensions[2], 2:3]
      }
      if(x_value %in% c(2:(dimensions[1]-1)) & y_value %in% c(2:(dimensions[2]-1))){
        neighbors[1,] <- schelling[x == x_value-1 & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == y_value-1, 2:3] 
      }
      if(x_value %in% c(2:(dimensions[1]-1)) & y_value == dimensions[2]){
        neighbors[1,] <- schelling[x == x_value-1 & y == 1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == 1, 2:3]
        neighbors[3,] <- schelling[x == x_value+1 & y == 1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == x_value+1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == x_value+1 & y == y_value-1, 2:3]
      }
      if(x_value == dimensions[1] & y_value == 1){
        neighbors[1,] <- schelling[x == x_value-1 & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == 1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == 1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == dimensions[2], 2:3]
        neighbors[7,] <- schelling[x == x_value & y == dimensions[2], 2:3]
        neighbors[8,] <- schelling[x == 1 & y == dimensions[2], 2:3]
      }
      if(x_value == dimensions[1] & y_value %in% c(2:(dimensions[2]-1))){
        neighbors[1,] <- schelling[x == x_value-1 & y == y_value+1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == y_value+1, 2:3]
        neighbors[3,] <- schelling[x == 1 & y == y_value+1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == 1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == 1 & y == y_value-1, 2:3]
      }
      if(x_value == dimensions[1] & y_value == dimensions[2]){
        neighbors[1,] <- schelling[x == x_value-1 & y == 1, 2:3]
        neighbors[2,] <- schelling[x == x_value & y == 1, 2:3]
        neighbors[3,] <- schelling[x == 1 & y == 1, 2:3]
        neighbors[4,] <- schelling[x == x_value-1 & y == y_value, 2:3]
        neighbors[5,] <- schelling[x == 1 & y == y_value, 2:3]
        neighbors[6,] <- schelling[x == x_value-1 & y == y_value-1, 2:3]
        neighbors[7,] <- schelling[x == x_value & y == y_value-1, 2:3]
        neighbors[8,] <- schelling[x == 1 & y == y_value-1, 2:3]
      }
      names(neighbors) <- c("x","y")
      
      ids <- schelling[x %in% neighbors$x & y %in% neighbors$y, id]
      
      return(ids)
    }
    # checks if the home is empty
    if (cur_race == "empty"){
      return(FALSE) # empty houses can't be unsatisfied
    }else{
      neighbor.ids <- select.neighbors(dimensions = dims)
      
      # counts simliar neighbors
      count_similar <- sum(schelling[id %in% neighbor.ids, 
                                     id != cur_id &
                                       race == cur_race])
      # counts different neighbors
      count_different <- sum(schelling[id %in% neighbor.ids, 
                                       id != cur_id &
                                         race != cur_race & 
                                         race != "empty"])
      
      # calculates the ratio
      ratio <- count_similar/(count_similar + count_different)
      
      # returns TRUE if the ratio is less than or equal to the threshold
      if(cur_race == races[1]){
        return(ratio <= dom_sim_threshold)
      }else{
        return(ratio <= min_sim_threshold)
      }
    }
  }
  
  # creates a ProgressBar
  pb <- txtProgressBar(min = 0, max = 1, style = 3)
  # and tracks time
  t <- Sys.time()
  
  #sets empty vector of ratios
  ratios.unsatisfied <<- c(NULL)
  
  # iterates
  for (iterate in 1:n){
    # fills the boolean vector "unsatisfied" 
    # indicates if the household is unsatisfied 
    schelling[, unsatisfied := is.unsatisfied_wrapped(x_value = x, 
                                                      y_value = y, 
                                                      dom_sim_threshold = dom_sim_threshold,
                                                      min_sim_threshold = min_sim_threshold),
              by = 1:nrow(schelling)]
    #hacky fix for NAs, which appear to happen when the ratio exactly equals the threshold (and should therefore evaluate to satisfied)
    schelling$unsatisfied[is.na(schelling$unsatisfied)] <- FALSE
    
    ratio.unsatisfied <- sum(schelling$unsatisfied)/nrow(schelling)
    ratios.unsatisfied <<- c(ratios.unsatisfied, ratio.unsatisfied)
    
    # move unsatisfied agents to an empty house
    # find the IDs that are empty where agents can migrate to
    emptyIDs <- schelling[race == "empty", id] # finds the id of empty houses
    
    usIDs <- schelling[(unsatisfied), id] # finds the origin of the agents moving
    
    availIDs <<- sort(c(emptyIDs, usIDs)) # finds available IDs for each household moving
    
    #subfunction to find available cells within specified distance limit
    find.available <- function(cell, avails, lim){
      distance <- c(NULL) #cast empty vector to fill with distances to available cells
      for(i in 1:length(avails)){
        #fill distances vector with absolute values of distance (to keep all values positive) between current cell and each available cell
        distance[i] <- abs(schelling[id == cell, x] - schelling[id == avails[i], x]) + abs(schelling[id == cell, y] - schelling[id == avails[i], y])
      }
      a <- as.data.frame(cbind(avails, distance)) #specify distance from current cell to each available cell
      options <- with(a, avails[distance > 0 & distance <= lim]) #pare down options to cells less than or equal to limited number of moves
      if(length(options >= 1)){
        target.id <- sample(options, size = 1)
      }else{target.id <- cell}
      availIDs <<- avails[-which(avails == target.id)]
      return(avails[which(avails == target.id)])
    }
    
    # cast a new data.table showing migration from origin_id to target-id
    transition <- data.table(origin = usIDs, 
                             oldRace = schelling[id %in% usIDs, race])
    transition[, target := find.available(cell = origin, avails = availIDs, lim = limit),
               by = 1:nrow(transition)]
    
    # move agents to the new homes
    schelling[id %in% transition$origin]$race = "empty"
    schelling[id %in% transition$target]$race = transition$oldRace
    
    # orders and reset schelling
    schelling <<- schelling[order(id)]
    
    # updates the ProgressBar
    setTxtProgressBar(pb, iterate/n)
  }
  close(pb)
  timedif <- Sys.time() - t
  
  # print out statistics for the calculation time
  print(paste0("Time for calculation in seconds: ", round(timedif, 3), " or: ",
               round(n / as.numeric(timedif), 3), " iterations per second"))
  print(ratios.unsatisfied)
  plotSchelling(title = paste0("Schelling Segregation Model after ", n, " Iterations"))
}
```


We can test out how each of the models performs under different specifications. First, for the initial model (unsatisfied agents move to random available cell with no distance limits), we plot the initial 20x20 wraparound Schelling universe with two races (divided 70/30) and 20 percent empty cells. We then plot the same universe after ten iterations assuming all agents prefer at least half of their neighbors to be of the same race. Finally, we plot the percentage of unsatisfied agents in this universe in each iteration.


```{r}
initiateSchelling_wrapped(dimensions = c(20,20), n_races = 2, perc_empty = .2, perc_maj = .7)
plotSchelling(title = "Schelling Segregation Model (Wrapped Environment)")
iterate_wrapped(n = 10, dom_sim_threshold = .5, min_sim_threshold = .5)
```

```{r}
#Plot pct satisfied over time
for(i in 1:length(ratios.unsatisfied)){
  if(i==1){
    plot(-100, -100, xlim=c(1,10), ylim=c(0,1), 
         ylab="Pct Unsatisfied", 
         xlab="Iteration", 
         main = "Percent of Agents Unsatisfied over Time",
         type="n", cex.axis=0.8)
  }else{
    segments(i-1, ratios.unsatisfied[i-1], i, ratios.unsatisfied[i], col="black", lwd=1)
  }
}
```


Here we see that the proportion of unsatisfied agents falls from 34 percent to just over 7 percent after ten rounds of movement.

Next, we see how these interactions change when we require that unsatisfied agents move to their nearest available cell. Again, we specify a 20x20 wraparound universe with two races (70/30 split) that both prefer at least 50 percent like neighbors, with 20 percent empty cells.


```{r}
initiateSchelling_wrapped(dimensions = c(20,20), n_races = 2, perc_empty = .2, perc_maj = .7)
plotSchelling(title = "Schelling Segregation Model (Wrapped Environment)")
iterate_wrapped_nearest(dom_sim_threshold = .5, min_sim_threshold = .5)
```



```{r}
#Plot pct unsatisfied over time
for(i in 1:length(ratios.unsatisfied)){
  if(i==1){
    plot(-100, -100, xlim=c(1,10), ylim=c(0,1), 
         ylab="Pct Unsatisfied", 
         xlab="Iteration", 
         main = "Percent of Agents Unsatisfied over Time",
         type="n", cex.axis=0.8)
  }else{
    segments(i-1, ratios.unsatisfied[i-1], i, ratios.unsatisfied[i], col="black", lwd=1)
  }
}
```


This doesn't change the general pattern. After ten iterations, the percent of unsatisfied agents has fallen similarly from just over 35 percent to just over 8 percent.

Finally, we see how the same Schelling universe and population behaves when they are freed up to move to any avaialble cell within specified distance limits. In this case, unsatisfied agents in the same population with the same preferences for like neighbors is limited to moving no more than six one-unit steps per round.


```{r}
initiateSchelling_wrapped(dimensions = c(20,20), n_races = 2, perc_empty = .2, perc_maj = .7)
plotSchelling(title = "Schelling Segregation Model (Wrapped Universe)")
iterate_wrapped_limit(dom_sim_threshold = .5, min_sim_threshold = .5, limit = 6)
```



```{r}
#Plot pct unsatisfied over time
for(i in 1:length(ratios.unsatisfied)){
  if(i==1){
    plot(-100, -100, xlim=c(1,10), ylim=c(0,1), 
         ylab="Pct Unsatisfied", 
         xlab="Iteration", 
         main = "Percent of Agents Unsatisfied over Time",
         type="n", cex.axis=0.8)
  }else{
    segments(i-1, ratios.unsatisfied[i-1], i, ratios.unsatisfied[i], col="black", lwd=1)
  }
}
```


This run started with a lower percent of unsatisfied agents (27.5%) and fell to roughly eight percent unsatisfied within seven moves -- not a major deviation from the pattern observed in the previous runs.

But what happens when we change other specifications in our model? Say, for instance, that the different races have different preferences for the number of like neighbors, given that they represent different shares of the overall population? We re-run the first model with those specifications below -- the majority race makes up 70 percent of the population and prefers to have at least 70 percent like neighbors; the minority race makes up 30 percent of the population and will move if less than 30 percent of their neighbors are of the same race:


```{r}
initiateSchelling_wrapped(dimensions = c(20,20), n_races = 2, perc_empty = .2, perc_maj = .7)
plotSchelling(title = "Schelling Segregation Model (Wrapped Universe)")
iterate_wrapped(dom_sim_threshold = .7, min_sim_threshold = .3)
```



```{r}
#Plot pct unsatisfied over time
for(i in 1:length(ratios.unsatisfied)){
  if(i==1){
    plot(-100, -100, xlim=c(1,10), ylim=c(0,1), 
         ylab="Pct Unsatisfied", 
         xlab="Iteration", 
         main = "Percent of Agents Unsatisfied over Time",
         type="n", cex.axis=0.8)
  }else{
    segments(i-1, ratios.unsatisfied[i-1], i, ratios.unsatisfied[i], col="black", lwd=1)
  }
}
```


This changes the pattern slightly, with the percent of unsatisfied agents falling from ~40% to ~6% in ten moves. Next, we'll test what happens when we change the number of races present and keep all of the other rules stable. In this case, specifying four races, with the majority race at 70% of the population and preferring at least 70% like neighbors and three minority races that each prefer to live with at least ten percent like neighbors (effectively, minority races only need one like neighbor in order to be satisfied):



```{r}
initiateSchelling_wrapped(dimensions = c(20,20), n_races = 4, perc_empty = .2, perc_maj = .7)
plotSchelling(title = "Schelling Segregation Model")
iterate_wrapped(dom_sim_threshold = .7, min_sim_threshold = .1)
```



```{r}
#Plot pct unsatisfied over time
for(i in 1:length(ratios.unsatisfied)){
  if(i==1){
    plot(-100, -100, xlim=c(1,10), ylim=c(0,1), 
         ylab="Pct Unsatisfied", 
         xlab="Iteration", 
         main = "Percent of Agents Unsatisfied over Time",
         type="n", cex.axis=0.8)
  }else{
    segments(i-1, ratios.unsatisfied[i-1], i, ratios.unsatisfied[i], col="black", lwd=1)
  }
}
```


Here we see a slightly different pattern. The percent of unsatisfied agents falls very quickly at first, going from just under 40 percent to just over 15 percent in five iterations. However, in the following five iterations the percent unsatisfied never falls below 14 percent, having seemingly reached an equilibrium more quickly than in previous runs.

Lastly, we look at what happens when the percent of empty cells changes, specifying a run with 5 percent empty cells and a run with 35 percent empty cells:


```{r}
initiateSchelling_wrapped(dimensions = c(20,20), n_races = 2, perc_empty = .05, perc_maj = .7)
plotSchelling(title = "Schelling")
iterate_wrapped(dom_sim_threshold = .5, min_sim_threshold = .5)
```



```{r}
#Plot pct unsatisfied over time
for(i in 1:length(ratios.unsatisfied)){
  if(i==1){
    plot(-100, -100, xlim=c(1,10), ylim=c(0,1), 
         ylab="Pct Unsatisfied", 
         xlab="Iteration", 
         main = "Percent of Agents Unsatisfied over Time",
         type="n", cex.axis=0.8)
  }else{
    segments(i-1, ratios.unsatisfied[i-1], i, ratios.unsatisfied[i], col="black", lwd=1)
  }
}

initiateSchelling_wrapped(dimensions = c(20,20), n_races = 2, perc_empty = .35, perc_maj = .7)
plotSchelling(title = "Schelling Segregation Model")
iterate_wrapped(dom_sim_threshold = .5, min_sim_threshold = .5)
```



```{r}
#Plot pct unsatisfied over time
for(i in 1:length(ratios.unsatisfied)){
  if(i==1){
    plot(-100, -100, xlim=c(1,10), ylim=c(0,1), 
         ylab="Pct Unsatisfied", 
         xlab="Iteration", 
         main = "Percent of Agents Unsatisfied over Time",
         type="n", cex.axis=0.8)
  }else{
    segments(i-1, ratios.unsatisfied[i-1], i, ratios.unsatisfied[i], col="black", lwd=1)
  }
}
```


With few empty cells to move to, movement by unsatisfied agents is restricted such that they primarily move to the cells previously occupied by other unsatisfied agents, reducing the likelihood that they will find a new cell with a higher number of like neighbors. As such, the percent of unsatisfied agents in the run with no empty cells does not fall as quickly as before -- going from just over 36 percent to just over 16 percent in ten iterations and having seemingly reached an equilibrium there. On the other hand, with more empty cells, agents are less likely to encounter non-like agents when they move, and as such the percent of unsatisfied agents falls from 28 percent to just under 4 percent in ten iterations.